Weight-adjusted discontinuous Galerkin methods: matrix-valued weights and elastic wave propagation in heterogeneous media

### 2017-10-02 docs/elasticity_notes/letter.tex

\begin{document}
\begin{letter}{Dr.\ Hans De Sterck\\SIAM J. Sci. Comp. Section Editor\\Methods and Algorithms for Scientific Computing}

\opening{Dear Dr.\ Hans De Sterck}

Please find enclosed the manuscript, 
\begin{center}
Jesse Chan\\
\textit{Weight adjusted discontinuous Galerkin methods: matrix-valued weights and elastic wave propagation in heterogeneous media},
\end{center}
which we would like to submit for publication as an original research article in the SIAM Journal on Scientific Computing.  This work extends weight-adjusted inner products, which provide a low-storage approximation of the inverse of a mass matrix with respect to an $L^2$ inner product with a spatially varying weight, to the case of spatially varying matrix-valued weights.  These approximations are then applied to the simulation of elastic wave propagation in arbitrary heterogeneous media, including anisotropy and media with sub-element variations.  We also introduce a new discontinuous Galerkin method which simplifies numerical fluxes while retaining energy stability and high order accuracy.

We hope that the method and results discussed in this manuscript will appeal to the readership of SIAM Journal on Scientific Computing.  We confirm that this manuscript has not been published elsewhere and is not under consideration by another journal.  We look forward to hearing from you at your earliest convenience.

>>>>

\begin{document}
\begin{letter}{Dr.\ Charbel Farhat\\Editor of IJNME}

\opening{Dear Dr.\ Farhat}

Please find enclosed the manuscript, 
\begin{center}
Jesse Chan\\
\textit{Weight adjusted discontinuous Galerkin methods: matrix-valued weights and elastic wave propagation in heterogeneous media},
\end{center}
which we would like to submit for publication as an original research article in IJNME.

This work extends weight-adjusted inner products, which provide a low-storage approximation of the inverse of a mass matrix with respect to an $L^2$ inner product with a spatially varying weight, to the case of spatially varying matrix-valued weights.  These approximations are then applied to the time-domain simulation of elastic wave propagation in arbitrary heterogeneous media, including anisotropy and media with sub-element variations.  The novelty of this work also includes a new energy stable and high order accurate discontinuous Galerkin method with simplified numerical fluxes for elastic wave propagation in arbitrary media.

We hope that the method and results discussed in this manuscript will appeal to the readership of IJNME.  We confirm that this manuscript has not been published elsewhere and is not under consideration by another journal.  We look forward to hearing from you at your earliest convenience.

### 2017-10-02 docs/elasticity_notes/main.tex

%\documentclass[preprint,10pt]{elsarticle}
%\documentclass[10pt]{article}
%\documentclass[review]{siamart0216}
\documentclass{siamart0216}

\usepackage{fullpage}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage[titletoc,toc,title]{appendix}

\usepackage{array} 
\usepackage{listings}
\usepackage{mathtools}
\usepackage{pdfpages}
\usepackage[textsize=footnotesize,color=green]{todonotes}
\usepackage{bm}
%\usepackage{tikz}
\usepackage[normalem]{ulem}
\usepackage{hhline}

%% ====================================== alg package
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage{algorithmicx}
\algblock{ParFor}{EndParFor}
% customising the new block
\algnewcommand\algorithmicparfor{\textbf{parfor}}
\algnewcommand\algorithmicpardo{\textbf{do}}
\algnewcommand\algorithmicendparfor{\textbf{end\ parfor}}
\algrenewtext{ParFor}[1]{\algorithmicparfor\ #1\ \algorithmicpardo}
\algrenewtext{EndParFor}{\algorithmicendparfor}
%% ====================================== end alg package

@@

%\newtheorem{theorem}{Theorem}[section]
%\newtheorem{lemma}[theorem]{Lemma}
%\newtheorem{proposition}[theorem]{Proposition}
%\newtheorem{corollary}[theorem]{Corollary}

@@

\newcommand{\note}[1]{{\color{blue}{#1}}}
\newcommand{\remark}[1]{\textbf{\color{red}#1}}

@@

@@

\begin{abstract}
Weight-adjusted inner products \cite{chan2016weight1,chan2016weight2} are easily invertible approximations to weighted $L^2$ inner products.  These approximations can be paired with a discontinuous Galerkin (DG) discretization to produce a time-domain method for wave propagation which is low storage, energy stable, and high order accurate for arbitrary heterogeneous media and curvilinear meshes.  In this work, we extend weight-adjusted DG (WADG) methods to the case of matrix-valued weights, with the linear elastic wave equation as an application.  We present a DG formulation of the symmetric form of the linear elastic wave equation, with upwind-like dissipation incorporated through simple penalty fluxes.  A semi-discrete convergence analysis is given, and numerical results confirm the stability and high order accuracy of WADG for several problems in elastic wave propagation.  
\end{abstract}

@@

\section{Symmetric form of the elastic wave equation}
\label{sec:symelas}
We begin with the linear elastic wave equation in a domain $\Omega \in \mathbb{R}^d$.  These equations can be written as a first order velocity-stress system for velocity $\bm{v}$ and symmetric stress tensor $\bm{S}$ 
\begin{align*}
\rho \pd{\bm{v}}{t} &= \Div{\bm{S}} + \bm{f}\\
\pd{\bm{S}}{t} &= \frac{1}{2} \tensor{\bm{C}}\LRp{\Grad\bm{v} + \Grad\bm{v}^T},
\end{align*}
where $\bm{f}$ is the body force per unit volume, $\rho$ is density, and $\tensor{\bm{C}}$ is the symmetric constitutive stiffness tensor relating stress and strain.  We rewrite these equations as a symmetric hyperbolic system of PDEs \cite{hughes1978classical} using Voigt notation
\begin{align}
\rho \pd{\bm{v}}{t} &= \sum_{i=1}^d \bm{A}_i^T \pd{\bm{\sigma}}{\bm{x}_i}\nonumber + \bm{f}\\
\bm{C}^{-1} \pd{\bm{\sigma}}{t} &= \sum_{i=1}^d \bm{A}_i \pd{\bm{v}}{\bm{x}_i}.
\label{eq:symelas}
\end{align}
where $\bm{C}$ is the symmetric matrix form of the constitutive tensor $\tensor{\bm{C}}$ and $\bm{\sigma}$ is a vector of length $N_d = \frac{d(d+1)}{2}$, the number of unique entries of the stress tensor $\bm{S}$ in $d$ dimensions.  We note that the matrices $\bm{A}_i$ are spatially constant, while $\rho,\bm{C}$, and $\bm{C}^{-1}$ can vary spatially.  Furthermore, we will assume that $\rho$ and $\bm{C}$ are positive-definite and bounded pointwise such that 
\begin{align*}
0 &< \rho_{\min} \leq \rho(\bm{x}) \leq \rho_{\max} < \infty, \\
0 &< c_{\min} \leq \bm{u}^T\bm{C}(\bm{x})\bm{u} \leq c_{\max} < \infty\\
0 &< \tilde{c}_{\min} \leq \bm{u}^T\bm{C}^{-1}(\bm{x})\bm{u} \leq \tilde{c}_{\max} < \infty
%, \qquad \forall \bm{u} \in \mathbb{R}^{N_d}, \quad \forall \bm{x} \in \mathbb{R}^d.
\end{align*}
for all $\bm{x}\in \mathbb{R}^d$ and all $\bm{u} \in \mathbb{R}^{N_d}$.

@@

One drawback of the analysis presented in this work is that accuracy of the weight-adjusted approximation is not guaranteed in the incompressible limit $\mu / \lambda \rightarrow 0$.  In this case, the stiffness matrix $\bm{C}$ becomes singular, and the constant $\tilde{w}_{\max}$ in the upper bound on $\nor{\bm{C}^{-1}}$ blows up.  However, numerical experiments in Section~\ref{sec:incomp} suggest that, while taking $\mu/\lambda \approx 0$ (or $\bm{C}$ nearly singular) results in larger relative errors for $\bm{\sigma}$, the accuracy of the WADG solution for $\bm{v}$ does not degrade significantly for near-incompressible materials.

>>>>

%\documentclass[preprint,10pt]{elsarticle}
\documentclass[10pt]{article}
%\documentclass[review]{siamart0216}
%\documentclass{siamart0216}

%\documentclass[AMA]{WileyNJD-v1}
%\articletype{Article Type}

\usepackage{fullpage}

\usepackage{lineno}

\usepackage{amsmath,amssymb,amsfonts}
\usepackage{amsthm}
\usepackage[titletoc,toc,title]{appendix}

\usepackage{array} 
\usepackage{listings}
\usepackage{mathtools}
%\usepackage{pdfpages}
%\usepackage[textsize=footnotesize,color=green]{todonotes}
\usepackage{bm}
\usepackage{tikz}
\usepackage[normalem]{ulem}
\usepackage{hhline}

@@

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}

%\newtheorem{proposition}[theorem]{Proposition}
%\newtheorem{corollary}[theorem]{Corollary}

@@

\newcommand{\note}[1]{{\color{blue}{#1}}}
%\newcommand{\remark}[1]{\textbf{\color{red}#1}}

@@

%\title{This is the sample article title\protect\thanks{This is an example for title footnote.}}
%\author[1]{Author One*}
%\authormark{AUTHOR ONE \textsc{et al}}
%\address[1]{\orgdiv{Original Division}, \orgname{Original University}, \orgaddress{\state{State name}, \country{Country name}}}
%\corres{*Corresponding author name, This is sample corresponding address. \email{authorone@gmail.com}}

@@

\linenumbers
\begin{abstract}
Weight-adjusted inner products \cite{chan2016weight1,chan2016weight2} are easily invertible approximations to weighted $L^2$ inner products.  These approximations can be paired with a discontinuous Galerkin (DG) discretization to produce a time-domain method for wave propagation which is low storage, energy stable, and high order accurate for arbitrary heterogeneous media and curvilinear meshes.  In this work, we extend weight-adjusted DG (WADG) methods to the case of matrix-valued weights, with the linear elastic wave equation as an application.  We present a DG formulation of the symmetric form of the linear elastic wave equation, with upwind-like dissipation incorporated through simple penalty fluxes.  A semi-discrete convergence analysis is given, and numerical results confirm the stability and high order accuracy of WADG for several problems in elastic wave propagation.  
\end{abstract}

@@

\section{Symmetric form of the elastic wave equation}
\label{sec:symelas}
We begin with the linear elastic wave equation in a domain $\Omega \in \mathbb{R}^d$.  These equations can be written as a first order velocity-stress system for velocity $\bm{v}$ and symmetric stress tensor $\bm{S}$ 
\begin{align*}
\rho \pd{\bm{v}}{t} &= \Div{\bm{S}} + \bm{f}\\
\pd{\bm{S}}{t} &= \frac{1}{2} \tensor{\bm{C}}\LRp{\Grad\bm{v} + \Grad\bm{v}^T},
\end{align*}
where $\bm{f}$ is the body force per unit volume, $\rho$ is density, and $\tensor{\bm{C}}$ is the symmetric constitutive stiffness tensor relating stress and strain.  We rewrite these equations as a symmetric hyperbolic system of PDEs \cite{hughes1978classical} using Voigt notation
\begin{align}
\rho \pd{\bm{v}}{t} &= \sum_{i=1}^d \bm{A}_i^T \pd{\bm{\sigma}}{\bm{x}_i}\nonumber + \bm{f}\\
\bm{C}^{-1} \pd{\bm{\sigma}}{t} &= \sum_{i=1}^d \bm{A}_i \pd{\bm{v}}{\bm{x}_i}.
\label{eq:symelas}
\end{align}
where $\bm{C}$ is the symmetric matrix form of the constitutive tensor $\tensor{\bm{C}}$ and $\bm{\sigma}$ is a vector of length $N_d = \frac{d(d+1)}{2}$, the number of unique entries of the stress tensor $\bm{S}$ in $d$ dimensions.  We note that the matrices $\bm{A}_i$ are spatially constant, while $\rho,\bm{C}$, and $\bm{C}^{-1}$ can vary spatially.  Furthermore, we will assume that $\rho$ and $\bm{C}$ are positive-definite and bounded pointwise such that 
\begin{align*}
0 &< \rho_{\min} \leq \rho(\bm{x}) \leq \rho_{\max} < \infty, \\
0 &< c_{\min} \leq \bm{u}^T\bm{C}(\bm{x})\bm{u} \leq c_{\max} < \infty\\
0 &< \tilde{c}_{\min} \leq \bm{u}^T\bm{C}^{-1}(\bm{x})\bm{u} \leq \tilde{c}_{\max} < \infty
\end{align*}
for all $\bm{x}\in \mathbb{R}^d$ and all $\bm{u} \in \mathbb{R}^{N_d}$.

@@

One drawback of the analysis presented in this work is that accuracy of the weight-adjusted approximation is not guaranteed in the incompressible limit $\mu / \lambda \rightarrow 0$.  In this case, the stiffness matrix $\bm{C}$ becomes singular, and the constant $\tilde{w}_{\max}$ in the upper bound on $\nor{\bm{C}^{-1}}$ blows up.  However, numerical experiments in Section~\ref{sec:incomp} suggest that, while taking $\mu/\lambda \approx 0$ (or $\bm{C}$ nearly singular) results in larger relative errors for $\bm{\sigma}$, the accuracy of the WADG solution for $\bm{v}$ does not degrade as significantly for near-incompressible materials.
